---
title: "2024 United States Presidential Election Analysis"
format:
  html:
    code-fold: true
    code-tools: true
    df-print: paged
    code-line-numbers: true
    toc: true
editor: visual
execute:
  warning: false
  message: false
  keep-md: true
  echo: false
---

```{r, results='hide'}
#| label: load-packages
library(tidyverse)
library(dplyr)
library(ggplot2)
library(ggthemes)
library(readr)
library(knitr)
```

# Mortality Rate Study

Mortality is known as the death rate, the measurement of the number of deaths within units of time relative to population size. There can be various factors that can influence this rate by either increasing it or decreasing it. The question becomes, "what could cause an increase or decrease of mortality rate?"

```{r, results='hide'}
#| label: load-data
gdp <- read_csv("csv/gdp.csv")
child_mortality <- read_csv("csv/child-mortality.csv")
poverty <- read_csv("csv/poverty.csv")
```

## How Does Poverty and GDP Per Capita Effect Mortality?

```{r, results='hide'}
#| label: tidy-data
gdp <- gdp |>
    rename(
        country = `Entity`,
        code = `Code`,
        year = `Year`,
        gdp_pc = `GDP per capita`,
    ) |> select(country:gdp_pc,-code) |> 
    arrange(year)

child_mortality <- child_mortality |>
    rename(
        country = `Entity`,
        code = `Code`,
        year = `Year`,
        mortality_rate = `Under-five mortality rate`,
    ) |> arrange(year) |> select(-code)

poverty <- poverty |>
    rename(
        country = `Entity`,
        code = `Code`,
        year = `Year`,
        poverty_per = `$2.15 a day - Share of population in poverty` #$2.15 a day - Share of population in poverty
    ) |> arrange(year) |> select(-code)

world_data <- merge(
    poverty |> filter(year >= 1950),
    child_mortality |> filter(year >= 1950),
    by=c("country","year")
    ) |> arrange(year)

world_data <- merge(
    world_data,
    gdp |> filter(year >= 1950),
    by=c("country","year")
) |> arrange(year)
```

```{r, results='hide'}
#| label: summary
world_data_summary <- world_data |>
    group_by(year) |>
    summarize(
        avg_poverty = round(mean(poverty_per),1),
        avg_gdp = round(mean(gdp_pc),1),
        avg_mortality = round(mean(mortality_rate),1),
    )

world_data_summary2 <- world_data |>
    group_by(country) |>
    summarize(
        avg_poverty = round(mean(poverty_per),1),
        avg_gdp = round(mean(gdp_pc),1),
        avg_mortality = round(mean(mortality_rate),1),
    )
```

```{r, results='hide'}
#| label: plot
ggplot(
    world_data_summary2,
    aes(y=avg_mortality,x=avg_poverty, color=avg_gdp/100)
) + geom_point() + 
    ggrepel::geom_label_repel(
        aes(label = country),
        size = 2.5,
        fill = NA,
        label.size = NA,
        color="black"
  ) +
    labs(
        title="Average World: Mortality Rate vs Poverty % (1963-2022)",
        subtitle="Mortality rate of children under 5 years old and with rate of people making  $2.15 a day.",
        caption="Source: https://ourworldindata.org",
        color="GDP Per Capita (x100)",
        x="% of Population in Poverty",
        y="Mortality Rate"
    ) +
    scale_color_gradient(low = "#007F7F", high = "#7FFFFF")
```

The graph as shown above compares the average global mortality rate and the average global poverty rate by plots on a graph. Not only that however, it also colors those points based off of the GDP per capita. Furthermore, each individual point represents a country whos mortality rate and poverty rate from the year 1963 to the year 2022 has been averaged. Overall, the plot does indicate that there is a correlation between poverty and gdp per capita with mortality. This could rationalized by saying that countries with lower income might not have the infrastructure or resources needed to sustain themselves. As when you look at the plot, the plots that are much darker (meaning lower GDP per capita) are plotted higher on the y-axis (meaning higher mortality rate) and are often plotted more to the right on the x-axis (meaning higher poverty rate).

## How has Mortality, Poverty, and GDP Per Capita Changed Over Time? Does It Also Prove There Is A Correlation Between Economics and Mortality?

```{r, results='hide'}
#| label: plot-2
ggplot(
    world_data_summary |> 
        rename(`Poverty` = avg_poverty, `GDP` = avg_gdp, `Mortality` = avg_mortality) |>
        pivot_longer(`Poverty`:`Mortality`, names_to="measurement",values_to="values"),
    aes(x=year,y=values,group=measurement, color=measurement)
) + geom_line() +
    geom_point(size=0.8,aes(shape=measurement)) +
    scale_x_log10(breaks = c(1963, 1970, 1980, 1990, 2000, 2010, 2022)) +
    scale_y_log10() +
    guides(shape = "none") +
    labs(
        title="Global Average : Mortality Rate vs Poverty % vs GDP Per Capita",
        subtitle="Mortality rate of children under 5 years old and with rate of people making $2.15 a day",
        caption="Source: https://ourworldindata.org",
        x="Year",
        y="",
        color="Measurement"
    )
```

The graph above illustrates the relationship between the global average of mortality rate, percentage of population in poverty, and GDP per capita. When you look at the lines of mortality and poverty, they are quite related with each other, often going up with each other, down with each other, or stagnant with each other. This indeed reaffirms the notion that poverty and income have correlation with mortality. This is also accompanied by the GDP line, where as, whenever it increases, mortality and poverty decrease, and whenever it decreases, mortality and poverty increase.